---
title: "MAE vs Variogram Score on FluSight Forecasts"
subtitle: "Comparing location-independent and cross-location scoring of 2025-26 flu season median forecasts"
author: "Claude (Anthropic)"
date: "2026-03-04"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7
)
knitr::opts_knit$set(root.dir = here::here())
```

## Overview

This analysis compares two ways of scoring FluSight median point forecasts for
the 2025-26 flu season:

- **MAE** (mean absolute error): scores each location independently, then
  averages. A model can have low MAE by being close in absolute terms at each
  location, even if it gets the relative pattern across locations wrong.
- **Variogram score**: evaluates whether a model captures the correct pairwise
  difference structure across locations. A model with perfect relative ordering
  and spacing gets a low variogram score even if it has a constant additive
  bias.

The goal is to see whether these two metrics agree on which models are "best",
and where they diverge.

**Requirements**: Dev versions of scoringutils (>= 2.1.2.9000) and hubEvals
(>= 0.1.0.9000) from GitHub. See the setup chunk below.

```{r install-dev-packages, eval=FALSE}
# Run once to install dev versions:
# remotes::install_github("epiforecasts/scoringutils")
# remotes::install_github("hubverse-org/hubEvals")
```

```{r libraries}
library(arrow)
library(hubData)
library(hubEvals)
library(scoringutils)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(DT)
```

## Load Forecasts from the Cloud

We connect to the FluSight hub's public S3 bucket, filter to the
`wk inc flu hosp` target with quantile output type for the 2025-26 season
(reference dates from 2025-11-22 through 2026-02-28), and keep horizons 0-3.

```{r load-forecasts}
hub_path <- s3_bucket("cdcepi-flusight-forecast-hub", anonymous = TRUE)

hub_con <- connect_hub(hub_path, file_format = "parquet", skip_checks = TRUE)

forecasts <- hub_con |>
  filter(
    target == "wk inc flu hosp",
    output_type == "quantile",
    horizon >= 0,
    reference_date >= "2025-11-22",
    reference_date <= "2026-02-28"
  ) |>
  collect_hub()

cat("Rows loaded:", nrow(forecasts), "\n")
cat("Models:", length(unique(forecasts$model_id)), "\n")
cat("Reference dates:", length(unique(forecasts$reference_date)), "\n")
```

### Select models with near-complete submissions

We count how many (reference_date, horizon, location) combos each model
submitted for, and keep models with >= 90% completeness.

```{r select-models}
ref_dates <- sort(unique(forecasts$reference_date))
horizons <- 0:3
locations <- sort(unique(forecasts$location))

# Count using just median forecasts (output_type_id == 0.5)
completeness <- forecasts |>
  filter(as.numeric(output_type_id) == 0.5) |>
  group_by(model_id) |>
  summarise(
    n_forecasts = n(),
    n_ref_dates = n_distinct(reference_date),
    .groups = "drop"
  )

n_possible <- length(ref_dates) * length(horizons) * length(locations)
completeness <- completeness |>
  mutate(pct_complete = n_forecasts / n_possible * 100) |>
  arrange(desc(pct_complete))

selected_models <- completeness |>
  filter(pct_complete >= 90) |>
  pull(model_id)

cat("Models with >= 90% completeness:", length(selected_models), "\n")

completeness |>
  filter(model_id %in% selected_models) |>
  mutate(pct_complete = round(pct_complete, 1)) |>
  datatable(
    colnames = c("Model", "Median forecasts", "Ref dates", "% complete"),
    options = list(pageLength = 15, order = list(list(3, "desc")))
  )
```

```{r filter-to-selected}
forecasts <- forecasts |>
  filter(model_id %in% selected_models)
```

## Compute MAE with hubEvals

We use `hubEvals::score_model_out()` with `metrics = "ae_median"` on the
quantile forecasts. This handles truth-joining internally via the oracle output.

```{r load-oracle}
oracle <- connect_target_oracle_output(hub_path) |>
  filter(target == "wk inc flu hosp") |>
  collect() |>
  select(-as_of)

cat("Oracle rows for wk inc flu hosp:", nrow(oracle), "\n")
```

```{r compute-mae}
mae_scores <- score_model_out(
  forecasts,
  oracle,
  metrics = "ae_median",
  summarize = FALSE
)

cat("MAE score rows:", nrow(mae_scores), "\n")
```

### Aggregate MAE

```{r mae-by-model}
mae_by_model <- mae_scores |>
  group_by(model_id) |>
  summarise(mae = mean(ae_median, na.rm = TRUE), .groups = "drop") |>
  arrange(mae)

mae_by_model |>
  mutate(mae = round(mae, 2)) |>
  datatable(
    colnames = c("Model", "Mean MAE"),
    options = list(pageLength = 15, order = list(list(1, "asc")))
  )
```

```{r mae-by-model-horizon}
mae_by_model_horizon <- mae_scores |>
  group_by(model_id, horizon) |>
  summarise(mae = mean(ae_median, na.rm = TRUE), .groups = "drop")
```

## Compute Variogram Score

The variogram score evaluates whether a model captures the correct pairwise
difference structure across locations. We use
`hubEvals::transform_quantile_model_out()` to join truth via the oracle output,
then extract median forecasts and pass them to
`scoringutils::as_forecast_multivariate_point()` with
`joint_across = "location"`. We exclude "US" (not independent of states).

### Prepare data via hubEvals

```{r prepare-vs-data}
# Use hubEvals to join truth to forecasts via oracle output
forecast_quantile <- transform_quantile_model_out(forecasts, oracle)

# Extract median rows (quantile_level == 0.5) and reshape for multivariate scoring
vs_data <- forecast_quantile |>
  as.data.frame() |>
  filter(quantile_level == 0.5, location != "US") |>
  select(model = model, reference_date, horizon, target_end_date,
         location, observed, predicted) |>
  rename(model_id = model) |>
  filter(!is.na(observed)) |>
  mutate(
    target_end_date = as.character(target_end_date),
    reference_date = as.character(reference_date)
  )

cat("VS data rows after truth join and filtering:", nrow(vs_data), "\n")
```

### Filter to complete snapshots

The variogram score requires all locations present for each
(model_id, reference_date, horizon) group.

```{r complete-snapshots}
n_locations <- n_distinct(vs_data$location)
cat("Expected locations per snapshot (excluding US):", n_locations, "\n")

snapshot_counts <- vs_data |>
  group_by(model_id, reference_date, horizon) |>
  summarise(n_loc = n_distinct(location), .groups = "drop")

incomplete <- snapshot_counts |> filter(n_loc < n_locations)
cat("Incomplete snapshots (will be dropped):", nrow(incomplete), "\n")

vs_data <- vs_data |>
  semi_join(
    snapshot_counts |> filter(n_loc == n_locations),
    by = c("model_id", "reference_date", "horizon")
  )

vs_data <- vs_data |> arrange(model_id, reference_date, horizon, location)
cat("VS data rows after filtering to complete snapshots:", nrow(vs_data), "\n")
```

### Score with scoringutils

```{r compute-vs}
vs_forecast <- as_forecast_multivariate_point(
  vs_data,
  forecast_unit = c("model_id", "reference_date", "horizon", "location"),
  joint_across = "location",
  observed = "observed",
  predicted = "predicted"
)

vs_scores <- score(vs_forecast)
cat("VS score rows:", nrow(vs_scores), "\n")
```

### Aggregate variogram scores

```{r vs-by-model}
vs_by_model <- vs_scores |>
  group_by(model_id) |>
  summarise(vs = mean(variogram_score, na.rm = TRUE), .groups = "drop") |>
  arrange(vs)

vs_by_model |>
  mutate(vs = round(vs, 1)) |>
  datatable(
    colnames = c("Model", "Mean VS"),
    options = list(pageLength = 15, order = list(list(1, "asc")))
  )
```

```{r vs-by-model-horizon}
vs_by_model_horizon <- vs_scores |>
  group_by(model_id, horizon) |>
  summarise(vs = mean(variogram_score, na.rm = TRUE), .groups = "drop")
```

## Compare MAE vs Variogram Score

### Combined rankings

```{r combined-table}
combined <- mae_by_model |>
  inner_join(vs_by_model, by = "model_id") |>
  inner_join(
    completeness |> select(model_id, pct_complete),
    by = "model_id"
  ) |>
  mutate(
    mae_rank = rank(mae),
    vs_rank = rank(vs)
  ) |>
  arrange(mae_rank)

rho <- cor(combined$mae_rank, combined$vs_rank, method = "spearman")

combined |>
  mutate(
    mae = round(mae, 2),
    vs = round(vs, 1),
    mae_rank = as.integer(mae_rank),
    vs_rank = as.integer(vs_rank),
    pct_complete = round(pct_complete, 1)
  ) |>
  datatable(
    colnames = c("Model", "MAE", "VS", "% complete", "MAE rank", "VS rank"),
    caption = sprintf("Spearman rank correlation: %.3f", rho),
    options = list(pageLength = 35, order = list(list(4, "asc")))
  )
```

### Plot 1: MAE by model, faceted by horizon

```{r plot-mae-horizon, fig.height=8}
model_order_mae <- mae_by_model |> arrange(mae) |> pull(model_id)

p_mae <- mae_by_model_horizon |>
  mutate(model_id = factor(model_id, levels = model_order_mae)) |>
  ggplot(aes(x = mae, y = model_id)) +
  geom_point(size = 2, color = "#2166AC") +
  facet_wrap(~paste("Horizon", horizon), scales = "free_x", nrow = 1) +
  labs(title = "MAE by model and horizon",
       x = "Mean Absolute Error", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

p_mae
```

### Plot 2: Variogram score by model, faceted by horizon

```{r plot-vs-horizon, fig.height=8}
model_order_vs <- vs_by_model |> arrange(vs) |> pull(model_id)

p_vs <- vs_by_model_horizon |>
  mutate(model_id = factor(model_id, levels = model_order_vs)) |>
  ggplot(aes(x = vs, y = model_id)) +
  geom_point(size = 2, color = "#B2182B") +
  facet_wrap(~paste("Horizon", horizon), scales = "free_x", nrow = 1) +
  labs(title = "Variogram score by model and horizon",
       x = "Variogram Score", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

p_vs
```

### Plot 3: MAE vs variogram score scatter (key plot)

```{r plot-scatter, fig.height=7, fig.width=8}
p_scatter <- combined |>
  ggplot(aes(x = mae, y = vs, label = model_id)) +
  geom_point(size = 3, color = "#4393C3") +
  geom_text_repel(size = 3, max.overlaps = 20, seed = 42) +
  labs(
    title = "MAE vs Variogram Score (overall means)",
    subtitle = sprintf("Spearman rank correlation: %.3f", rho),
    x = "Mean Absolute Error (lower = better)",
    y = "Variogram Score (lower = better)"
  ) +
  theme_minimal(base_size = 12)

p_scatter
ggsave("variogram-flusight-analysis/plots/mae_vs_vs_scatter.png",
       p_scatter, width = 9, height = 7, dpi = 150)
```

### Plot 4: Rank comparison

```{r plot-ranks, fig.height=7, fig.width=7}
p_ranks <- combined |>
  ggplot(aes(x = mae_rank, y = vs_rank, label = model_id)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, color = "#4393C3") +
  geom_text_repel(size = 3, max.overlaps = 20, seed = 42) +
  labs(
    title = "Rank comparison: MAE rank vs Variogram Score rank",
    subtitle = "Points on diagonal = same rank under both metrics",
    x = "MAE rank (1 = best)",
    y = "Variogram Score rank (1 = best)"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

p_ranks
ggsave("variogram-flusight-analysis/plots/rank_comparison.png",
       p_ranks, width = 7, height = 7, dpi = 150)
```

### Plot 5: Variogram score over time (horizon 1)

```{r plot-vs-time, fig.height=6, fig.width=10}
highlight_models <- unique(c(
  head(model_order_vs, 3),
  tail(model_order_vs, 2),
  head(model_order_mae, 2)
))

vs_time <- vs_scores |>
  filter(horizon == 1) |>
  mutate(reference_date = as.Date(reference_date))

p_time <- vs_time |>
  ggplot(aes(x = reference_date, y = variogram_score,
             group = model_id)) +
  geom_line(data = vs_time |> filter(!model_id %in% highlight_models),
            alpha = 0.15, color = "grey50") +
  geom_line(data = vs_time |> filter(model_id %in% highlight_models),
            aes(color = model_id), linewidth = 0.8) +
  labs(
    title = "Variogram score over the season (horizon 1)",
    x = "Reference date",
    y = "Variogram Score",
    color = "Model"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_time
ggsave("variogram-flusight-analysis/plots/vs_over_time_h1.png",
       p_time, width = 10, height = 6, dpi = 150)
```

## Case Studies: Why MAE and VS Disagree

To find clear examples of MAE/VS disagreement, we first identify a single
reference date where the spread of both scores across models is large. We then
pick model pairs based on their rankings **on that specific date**, so the
snapshot plots directly reflect the chosen contrast.

```{r find-focus-date}
# Compute per-model, per-date scores at horizon 1
mae_h1 <- mae_scores |>
  filter(horizon == 1) |>
  group_by(model_id, reference_date) |>
  summarise(mae = mean(ae_median, na.rm = TRUE), .groups = "drop")

vs_h1 <- vs_scores |>
  filter(horizon == 1) |>
  mutate(reference_date = as.Date(reference_date))

# For each date, compute IQR of MAE and VS across models, then normalise
date_spread <- mae_h1 |>
  group_by(reference_date) |>
  summarise(mae_iqr = IQR(mae), .groups = "drop") |>
  inner_join(
    vs_h1 |>
      group_by(reference_date) |>
      summarise(vs_iqr = IQR(variogram_score), .groups = "drop"),
    by = "reference_date"
  ) |>
  mutate(
    mae_iqr_norm = mae_iqr / max(mae_iqr),
    vs_iqr_norm = vs_iqr / max(vs_iqr),
    combined_spread = mae_iqr_norm + vs_iqr_norm
  ) |>
  arrange(desc(combined_spread))

focus_date <- date_spread$reference_date[1]
cat("Focus date (largest combined MAE + VS spread at horizon 1):", as.character(focus_date), "\n")
```

```{r pick-case-models}
# Rank models on the focus date
mae_focus <- mae_h1 |>
  filter(reference_date == focus_date) |>
  mutate(mae_rank = rank(mae))

vs_focus <- vs_h1 |>
  filter(reference_date == focus_date) |>
  mutate(vs_rank = rank(variogram_score))

focus_ranks <- mae_focus |>
  select(model_id, mae, mae_rank) |>
  inner_join(
    vs_focus |> select(model_id, vs = variogram_score, vs_rank),
    by = "model_id"
  )

# Pick model pairs from the focus-date rankings
# Good at both: lowest sum of ranks
focus_ranks <- focus_ranks |>
  mutate(rank_sum = mae_rank + vs_rank,
         rank_diff = mae_rank - vs_rank)

good_both <- focus_ranks |>
  arrange(rank_sum) |>
  head(2) |>
  pull(model_id)

# Good MAE / bad VS: most negative rank_diff (low MAE rank, high VS rank)
good_mae_bad_vs <- focus_ranks |>
  arrange(rank_diff) |>
  head(2) |>
  pull(model_id)

# Good VS / bad MAE: most positive rank_diff (high MAE rank, low VS rank)
good_vs_bad_mae <- focus_ranks |>
  arrange(desc(rank_diff)) |>
  head(2) |>
  pull(model_id)

case_models <- c(good_both, good_mae_bad_vs, good_vs_bad_mae)

# Label each model with its group for faceting
case_labels <- tibble(
  model_id = case_models,
  group = factor(
    rep(c("Good both", "Good MAE / Bad VS", "Good VS / Bad MAE"), each = 2),
    levels = c("Good both", "Good MAE / Bad VS", "Good VS / Bad MAE")
  )
)

focus_ranks |>
  filter(model_id %in% case_models) |>
  inner_join(case_labels, by = "model_id") |>
  mutate(
    mae = round(mae, 1), vs = round(vs, 0),
    mae_rank = as.integer(mae_rank), vs_rank = as.integer(vs_rank)
  ) |>
  arrange(group) |>
  select(group, model_id, mae, mae_rank, vs, vs_rank) |>
  knitr::kable(col.names = c("Group", "Model", "MAE", "MAE rank", "VS",
                               "VS rank"),
               caption = sprintf("Model pairs selected from %s (horizon 1)", focus_date))
```

```{r load-locations}
# Load location abbreviations from S3 (no hubData helper for auxiliary data)
s3 <- S3FileSystem$create(anonymous = TRUE)
loc_lookup <- read_csv_arrow(
  s3$OpenInputStream(
    "cdcepi-flusight-forecast-hub/auxiliary-data/locations.csv"
  )
) |>
  as.data.frame() |>
  select(location, abbreviation)
```

### Decomposing the variogram score

The variogram score (order $p = 0.5$) is defined as:

$$\text{VS} = \sum_{i < j} \left( |y_i - y_j|^{0.5} - |\widehat{y}_i - \widehat{y}_j|^{0.5} \right)^2$$

where $y_i$ and $\widehat{y}_i$ are the observed and predicted values at location
$i$. The score sums over all $\binom{52}{2} = 1326$ location pairs. Each term
measures how well the model captures the *difference* between two locations.

To see what drives VS, write $\widehat{y}_i = y_i + e_i$ where $e_i$ is the
forecast error at location $i$. The predicted pairwise difference becomes
$\widehat{y}_i - \widehat{y}_j = (y_i - y_j) + (e_i - e_j)$. So each VS term depends
on the **error differential** $e_i - e_j$, not the errors themselves:

- If all errors are equal ($e_i = e_j$ for all pairs), the error differentials
  are zero and VS is perfect --- even with large constant bias.
- If one location has a very different error from the rest, all pairs involving
  it have large error differentials, inflating VS.

In other words, VS measures **error heterogeneity** across locations. A model
with uniformly large errors (bad MAE) can still get a good VS because the
pairwise differentials are small. A model with mostly small errors but one
outlier location (decent MAE) can get a bad VS because that one outlier creates
51 pairs with large differentials.

We can decompose each model's total VS into **per-location contributions** by
summing all pairwise terms that involve each location (each term is counted
twice, once per participating location). Locations whose error $e_i$ diverges
most from the other locations' errors will dominate the sum.

```{r vs-decomposition, fig.height=9, fig.width=10}
# Compute pairwise VS terms for a single snapshot (one model, one date)
compute_loc_contributions <- function(df) {
  n <- nrow(df)
  pairs <- expand.grid(i = 1:n, j = 1:n) |> filter(i < j)
  obs_diff <- abs(df$observed[pairs$i] - df$observed[pairs$j])^0.5
  pred_diff <- abs(df$predicted[pairs$i] - df$predicted[pairs$j])^0.5
  term <- (obs_diff - pred_diff)^2

  # Attribute each term to both participating locations
  pair_data <- data.frame(
    location = c(df$location[pairs$i], df$location[pairs$j]),
    vs_term = rep(term, 2)
  )
  pair_data |>
    group_by(location) |>
    summarise(vs_contribution = sum(vs_term), .groups = "drop")
}

# Compute for all case models on focus date, horizon 1
focus_snap <- vs_data |>
  filter(model_id %in% case_models,
         reference_date == as.character(focus_date), horizon == 1)

loc_vs <- focus_snap |>
  group_by(model_id) |>
  group_modify(~ compute_loc_contributions(.x)) |>
  ungroup() |>
  inner_join(case_labels, by = "model_id") |>
  group_by(model_id) |>
  mutate(
    pct_of_total = vs_contribution / sum(vs_contribution) * 100,
    loc_rank = rank(-vs_contribution)
  ) |>
  ungroup()

# Show cumulative contribution curves (Lorenz-like)
cum_curves <- loc_vs |>
  arrange(model_id, desc(vs_contribution)) |>
  group_by(model_id, group) |>
  mutate(
    cum_pct = cumsum(vs_contribution) / sum(vs_contribution) * 100,
    loc_index = row_number()
  ) |>
  ungroup() |>
  mutate(
    facet = factor(
      paste0(group, ": ", model_id),
      levels = loc_vs |>
        distinct(group, model_id) |>
        arrange(group, model_id) |>
        mutate(f = paste0(group, ": ", model_id)) |>
        pull(f)
    )
  )

p_cum <- cum_curves |>
  ggplot(aes(x = loc_index, y = cum_pct, color = group)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 50, linetype = "dotted", color = "grey50") +
  facet_wrap(~facet, ncol = 2) +
  scale_color_manual(values = c(
    "Good both" = "#2166AC",
    "Good MAE / Bad VS" = "#B2182B",
    "Good VS / Bad MAE" = "#1B7837"
  )) +
  labs(
    title = sprintf("Cumulative VS contribution by location rank (%s, horizon 1)", focus_date),
    subtitle = "Steeper curves = VS dominated by a few outlier locations; flatter = error differentials spread evenly",
    x = "Location rank (by VS contribution, highest first)",
    y = "Cumulative % of total VS"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"))

p_cum
ggsave("variogram-flusight-analysis/plots/vs_cumulative_contribution.png",
       p_cum, width = 10, height = 9, dpi = 150)
```

```{r vs-concentration-table}
# Summary: how concentrated is each model's VS?
concentration <- loc_vs |>
  group_by(model_id, group) |>
  summarise(
    total_vs = sum(vs_contribution) / 2,  # divide by 2 since each term counted twice
    top1_pct = max(pct_of_total),
    top5_pct = sum(sort(pct_of_total, decreasing = TRUE)[1:5]),
    .groups = "drop"
  ) |>
  arrange(group)

# Get the top location for each model
top_locs <- loc_vs |>
  group_by(model_id) |>
  slice_max(vs_contribution, n = 1) |>
  left_join(loc_lookup, by = "location") |>
  select(model_id, top_location = abbreviation, top_loc_pct = pct_of_total)

concentration |>
  left_join(top_locs, by = "model_id") |>
  mutate(
    total_vs = round(total_vs, 0),
    top1_pct = round(top1_pct, 1),
    top5_pct = round(top5_pct, 1),
    top_loc_pct = round(top_loc_pct, 1)
  ) |>
  select(group, model_id, total_vs, top_location, top1_pct, top5_pct) |>
  knitr::kable(
    col.names = c("Group", "Model", "Total VS", "Top location",
                  "Top-1 loc %", "Top-5 locs %"),
    caption = "Concentration of VS contributions across locations"
  )
```

The "good MAE / bad VS" models have VS contributions heavily concentrated in a
few locations --- these are the locations whose errors differ most from the rest,
creating large pairwise error differentials. In contrast, the "good VS / bad MAE"
models have contributions spread more evenly: their errors are large everywhere
(bad MAE), but roughly *uniform* across locations, so the pairwise differentials
stay small (good VS).

### Deep dive: UVAFluX-Ensemble's New York overshoot

UVAFluX-Ensemble provides a concrete illustration. On the focus date, it
dramatically overshoots New York ($e_{\text{NY}} \approx +5500$) while
undershooting most other large states ($e_j < 0$). Because NY's error diverges
so sharply from the rest, all 51 pairs involving NY have large error
differentials $e_{\text{NY}} - e_j$, inflating the VS far beyond what the MAE
would suggest.

```{r uva-deep-dive, fig.height=5, fig.width=10}
# Get the top contributing location for UVAFluX-Ensemble
uva_top_loc <- loc_vs |>
  filter(model_id == "UVAFluX-Ensemble") |>
  slice_max(vs_contribution, n = 1) |>
  pull(location)

uva_snap <- focus_snap |>
  filter(model_id == "UVAFluX-Ensemble") |>
  left_join(loc_lookup, by = "location") |>
  mutate(
    error = predicted - observed,
    is_outlier = location == uva_top_loc
  )

uva_top_abbr <- uva_snap$abbreviation[uva_snap$location == uva_top_loc]
uva_top_obs <- uva_snap$observed[uva_snap$location == uva_top_loc]
uva_top_pred <- uva_snap$predicted[uva_snap$location == uva_top_loc]
uva_top_err <- uva_top_pred - uva_top_obs

# Compare with a good-VS model on the same date
good_model <- good_both[1]
good_snap <- focus_snap |>
  filter(model_id == good_model) |>
  left_join(loc_lookup, by = "location") |>
  mutate(error = predicted - observed)

# Get dynamic ranks for panel labels
uva_ranks <- focus_ranks |> filter(model_id == "UVAFluX-Ensemble")
good_ranks <- focus_ranks |> filter(model_id == good_model)

# Side-by-side: error by location for UVA vs the good model
compare_data <- bind_rows(
  uva_snap |> mutate(panel = sprintf("UVAFluX-Ensemble (VS rank %d, MAE rank %d)",
                                      as.integer(uva_ranks$vs_rank),
                                      as.integer(uva_ranks$mae_rank))),
  good_snap |> mutate(panel = sprintf("%s (VS rank %d, MAE rank %d)",
                                       good_model,
                                       as.integer(good_ranks$vs_rank),
                                       as.integer(good_ranks$mae_rank)),
                       is_outlier = FALSE)
) |>
  mutate(panel = factor(panel, levels = unique(panel)))

p_compare <- compare_data |>
  ggplot(aes(x = reorder(abbreviation, -abs(error)), y = error,
             fill = is_outlier)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "#4393C3", "TRUE" = "#B2182B")) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(
    title = sprintf("Forecast errors by location (%s, horizon 1)", focus_date),
    subtitle = sprintf(
      "UVA overshoots %s by %s (obs: %s, pred: %s) --- this one location drives %.0f%% of UVA's total VS",
      uva_top_abbr, format(uva_top_err, big.mark = ","),
      format(uva_top_obs, big.mark = ","), format(uva_top_pred, big.mark = ","),
      loc_vs |> filter(model_id == "UVAFluX-Ensemble", location == uva_top_loc) |> pull(pct_of_total)
    ),
    x = "State (ordered by |error|)", y = "Error (predicted - observed)"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7),
        strip.text = element_text(face = "bold"))

p_compare
ggsave("variogram-flusight-analysis/plots/uva_error_comparison.png",
       p_compare, width = 10, height = 7, dpi = 150)
```

We can visualise the VS matrix directly. Each cell $(i, j)$ in the lower
triangle shows the pairwise VS term
$(|y_i - y_j|^{0.5} - |\widehat{y}_i - \widehat{y}_j|^{0.5})^2$.
Locations are sorted by forecast error so that the outlier location appears at
one edge. For UVA, the entire row/column corresponding to New York lights up;
for the good model, no single row or column dominates.

```{r vs-heatmap, fig.height=7, fig.width=14}
# Compute the full symmetric pairwise VS matrix for one model
compute_vs_matrix <- function(df) {
  # df must have: abbreviation, observed, predicted — sorted by error
  n <- nrow(df)
  pairs <- expand.grid(i = 1:n, j = 1:n) |> filter(i != j)
  obs_diff <- abs(df$observed[pairs$i] - df$observed[pairs$j])^0.5
  pred_diff <- abs(df$predicted[pairs$i] - df$predicted[pairs$j])^0.5
  term <- (obs_diff - pred_diff)^2
  data.frame(
    loc_i = df$abbreviation[pairs$i],
    loc_j = df$abbreviation[pairs$j],
    vs_term = term
  )
}

# Build one heatmap using a shared location order
make_heat_plot <- function(snap_df, model_label, loc_order) {
  snap_df <- snap_df |> arrange(match(abbreviation, loc_order))
  # Ensure abbreviation column exists and matches loc_order
  snap_df <- snap_df |> filter(abbreviation %in% loc_order)
  mat <- compute_vs_matrix(snap_df) |>
    mutate(
      loc_i = factor(loc_i, levels = loc_order),
      loc_j = factor(loc_j, levels = loc_order)
    )
  total <- format(round(sum(mat$vs_term) / 2), big.mark = ",")
  ggplot(mat, aes(x = loc_j, y = loc_i, fill = vs_term)) +
    geom_tile() +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "VS term",
                         trans = "sqrt", limits = shared_limits) +
    scale_y_discrete(limits = rev(loc_order)) +
    scale_x_discrete(position = "top") +
    labs(title = sprintf("%s (total VS = %s)", model_label, total),
         x = NULL, y = NULL) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 5),
      axis.text.y = element_text(size = 5),
      plot.title = element_text(face = "bold", size = 10)
    ) +
    coord_equal()
}

# Shared location order: UVA's per-location VS contribution (largest at top-left)
uva_loc_contrib <- loc_vs |>
  filter(model_id == "UVAFluX-Ensemble") |>
  left_join(loc_lookup, by = "location") |>
  arrange(desc(vs_contribution))
shared_loc_order <- uva_loc_contrib$abbreviation

# Compute shared colour limits across both models
uva_ordered <- uva_snap |> arrange(match(abbreviation, shared_loc_order))
good_ordered <- good_snap |> arrange(match(abbreviation, shared_loc_order))
all_terms <- c(compute_vs_matrix(uva_ordered)$vs_term,
               compute_vs_matrix(good_ordered)$vs_term)
shared_limits <- c(0, max(all_terms))

p_heat_uva <- make_heat_plot(uva_snap, "UVAFluX-Ensemble", shared_loc_order)
p_heat_good <- make_heat_plot(good_snap, good_model, shared_loc_order)

p_heat <- (p_heat_uva + theme(legend.position = "bottom")) |
  (p_heat_good + theme(legend.position = "bottom")) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = sprintf("Pairwise VS term matrices (%s, horizon 1)", focus_date),
    subtitle = "Locations sorted by UVA's per-location VS contribution (largest at top-left). Each cell = one term in the VS sum.",
    theme = theme(
      plot.title = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(size = 10)
    )
  )

p_heat
ggsave("variogram-flusight-analysis/plots/vs_pairwise_heatmap.png",
       p_heat, width = 14, height = 7, dpi = 150)
```

The UVA heatmap makes the mechanism visually obvious: the top row and leftmost
column (New York, the location with the largest error) are uniformly hot,
because the large error differential $e_{\text{NY}} - e_j$ produces a large VS
term for every partner location $j$. The good model's heatmap shows no such
stripe --- its VS terms are smaller overall and distributed more evenly across
the matrix.

### Cross-location profiles on the focus date

For the top 20 locations by observed value, we compare the predicted profile
to observed. Models with good VS should track the *shape* of the observed
profile, even if the level is off.

```{r case-snapshot-profile, fig.height=12, fig.width=10}
snapshot <- vs_data |>
  filter(model_id %in% case_models, horizon == 1,
         reference_date == as.character(focus_date)) |>
  left_join(loc_lookup, by = "location") |>
  inner_join(case_labels, by = "model_id")

# Order facets by group then model
facet_order <- snapshot |>
  distinct(group, model_id) |>
  arrange(group, model_id) |>
  mutate(f = paste0(group, ": ", model_id)) |>
  pull(f)

# Top 20 locations by observed value (from truth for this week)
top_locs <- snapshot |>
  filter(model_id == case_models[1]) |>
  arrange(desc(observed)) |>
  head(20) |>
  pull(abbreviation)

snapshot_top <- snapshot |>
  filter(abbreviation %in% top_locs) |>
  tidyr::pivot_longer(cols = c(observed, predicted),
                      names_to = "type", values_to = "value") |>
  mutate(
    abbreviation = factor(abbreviation, levels = top_locs),
    type = factor(type, levels = c("observed", "predicted")),
    facet = factor(
      paste0(group, ": ", model_id),
      levels = facet_order
    )
  )

p_snapshot <- snapshot_top |>
  ggplot(aes(x = abbreviation, y = value, fill = type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  scale_fill_manual(values = c("observed" = "grey40", "predicted" = "#4393C3")) +
  facet_wrap(~facet, scales = "free_y", ncol = 2) +
  labs(
    title = sprintf("Forecast vs observed for top 20 locations (%s, horizon 1)", focus_date),
    subtitle = "Rows: good both, good MAE / bad VS, good VS / bad MAE",
    x = "State (ordered by observed)", y = "Hospitalizations", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "bottom")

p_snapshot
ggsave("variogram-flusight-analysis/plots/snapshot_profile.png",
       p_snapshot, width = 10, height = 12, dpi = 150)
```

## Discussion

```{r summary-stats}
cat(sprintf("Spearman rank correlation (MAE vs VS): %.3f\n", rho))

combined |>
  mutate(
    rank_diff = abs(mae_rank - vs_rank),
    mae = round(mae, 2),
    vs = round(vs, 1),
    mae_rank = as.integer(mae_rank),
    vs_rank = as.integer(vs_rank)
  ) |>
  arrange(desc(rank_diff)) |>
  head(5) |>
  select(model_id, mae, vs, mae_rank, vs_rank, rank_diff) |>
  datatable(
    colnames = c("Model", "MAE", "VS", "MAE rank", "VS rank", "Rank diff"),
    caption = "Models with largest rank disagreement between MAE and VS",
    options = list(pageLength = 5, dom = "t")
  )
```

**Key observations**:

1. **Correlation between metrics**: A Spearman correlation of `r round(rho, 3)`
   indicates how much MAE and VS agree on model rankings. High correlation
   suggests most models that are good at pointwise accuracy also capture
   cross-location structure well. Low correlation would mean these are measuring
   genuinely different aspects of forecast quality.

2. **VS blindness to additive bias**: The variogram score is invariant to
   constant additive bias --- a model that overshoots every location by the same
   amount gets a perfect variogram score. Models that rank better on VS than MAE
   may have this property: good relative structure but a level shift.

3. **Horizon effects**: Both metrics typically degrade at longer horizons, but
   the rate of degradation may differ. If VS degrades more slowly, it suggests
   models maintain relative cross-location structure better than absolute
   accuracy at longer lead times.

## Session Info

```{r session-info}
sessionInfo()
```
