---
title: "Forecast scoring through small examples"
subtitle: "NYC ILI ED visits, 2025/26 season — illustrating how different scoring rules rank forecasts"
author: "Nicholas Reich"
date: "2026-05-13"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    fig-width: 11
    fig-height: 5.5
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
---

```{=html}
<style>
  /* Compact, top-aligned panel tabsets */
  .reveal .panel-tabset { margin-top: 0; }
  .reveal .panel-tabset > .nav-tabs {
    font-size: 0.5em;
    margin: 0 0 0.2em 0;
    padding: 0;
    border-bottom: 1px solid #999;
  }
  .reveal .panel-tabset > .nav-tabs .nav-link {
    padding: 0.1em 0.6em;
    margin: 0;
  }
  .reveal .panel-tabset > .tab-content {
    padding-top: 0.1em;
  }
</style>
```

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(readr)
library(scoringutils)
library(knitr)
```

## Setup

Each example uses real weekly observations from the [MetroCast Hub](https://github.com/reichlab/flu-metrocast) for **NYC, ILI ED visits %** during the 2025/26 season. Two stylised forecasts (Model X and Model Y) are shown with **95% prediction intervals**, then evaluated under several scoring rules.

The forecasts are illustrative caricatures designed to expose how each scoring rule responds to specific failure modes (bias, sharpness, joint coherence, etc.). They are not real submissions to the hub.

```{r load-data}
# latest-data.csv has full historical observed values (no as_of versioning);
# oracle-output.csv only covers the 2025/26 forecast/scoring window.
truth_url <- "https://raw.githubusercontent.com/reichlab/flu-metrocast/main/target-data/latest-data.csv"
truth <- read_csv(truth_url, show_col_types = FALSE) |>
  filter(location == "nyc", target == "ILI ED visits pct") |>
  mutate(target_end_date = as.Date(target_end_date)) |>
  rename(oracle_value = observation) |>
  arrange(target_end_date)
```

```{r forecast-helpers}
# Generate N_SAMPLES sample trajectories with prescribed per-horizon median and SD.
# Most of the variance is "shared across horizons" (a smooth per-sample offset)
# so trajectories look like smooth-ish curves rather than jagged noise.
# `dist = "normal"` gives symmetric intervals; `dist = "lognormal"` gives
# tighter lower tail and wider upper tail (natural for count-valued epi targets).
generate_forecast <- function(model_id, medians, sds, target_dates,
                              dist = c("normal", "lognormal"),
                              floor = 0,
                              n_samples = 200, shared_frac = 0.9, seed = NULL) {
  dist <- match.arg(dist)
  if (!is.null(seed)) set.seed(seed)
  shift_sd <- shared_frac * sds
  noise_sd <- sqrt(1 - shared_frac^2) * sds
  H <- length(medians)
  u <- rnorm(n_samples)   # per-sample standardized offset shared across horizons
  mat <- matrix(NA_real_, nrow = n_samples, ncol = H)
  for (h in seq_len(H)) {
    if (dist == "normal") {
      mat[, h] <- medians[h] + u * shift_sd[h] +
        rnorm(n_samples, 0, noise_sd[h])
    } else {
      # (Shifted) lognormal: value = floor + exp(mu + sigma*Z), with mu chosen
      # so that median(value) = medians[h] and sigma controls the upper-tail
      # spread. `floor` (>= 0) sets a soft lower bound on sample trajectories.
      stopifnot(all(medians > floor))
      log_val <- log(medians[h] - floor) + u * shift_sd[h] +
        rnorm(n_samples, 0, noise_sd[h])
      mat[, h] <- floor + exp(log_val)
    }
  }
  mat <- pmax(mat, 0)
  tibble(
    model_id = model_id,
    sample_id = rep(seq_len(n_samples), times = H),
    target_end_date = rep(target_dates, each = n_samples),
    horizon = rep(seq_len(H), each = n_samples),
    value = as.vector(mat)
  )
}

PI_LEVEL <- 0.95   # central PI level used throughout
PI_TAIL  <- (1 - PI_LEVEL) / 2  # e.g., 0.025 for a 95% PI

summarise_bands <- function(forecast_df) {
  forecast_df |>
    group_by(model_id, target_end_date, horizon) |>
    summarise(
      lower  = quantile(value, PI_TAIL,     type = 7),
      median = quantile(value, 0.50,        type = 7),
      upper  = quantile(value, 1 - PI_TAIL, type = 7),
      .groups = "drop"
    )
}

# Stacked plot, one of two scale modes:
#   scale = "linear" -> trajectories + median (top), 95% PI + median (bottom), linear y
#   scale = "log"    -> trajectories + median (top), 95% PI + median (bottom), log10 y
# Used inside Quarto panel-tabsets so the viewer can toggle scales.
plot_forecast_panel <- function(truth_window, forecast_df, ref_date,
                                title, subtitle = NULL,
                                model_colors = c("Model X" = "#4477AA",
                                                 "Model Y" = "#DDAA33"),
                                n_display_trajectories = 80,
                                y_lim = c(0, 16),
                                y_lim_log = c(0.8, 40),
                                scale = c("linear", "log"),
                                display_seed = 42) {
  scale <- match.arg(scale)
  bands <- summarise_bands(forecast_df)

  # Deterministic subsample so the linear and log tabs show the same trajectories
  set.seed(display_seed)
  display_ids <- forecast_df |>
    distinct(model_id, sample_id) |>
    group_by(model_id) |>
    slice_sample(n = n_display_trajectories) |>
    ungroup()
  forecast_display <- forecast_df |>
    semi_join(display_ids, by = c("model_id", "sample_id"))

  base_layers <- list(
    geom_line(data = truth_window,
              aes(x = target_end_date, y = oracle_value),
              colour = "grey20", linewidth = 0.7),
    geom_point(data = truth_window,
               aes(x = target_end_date, y = oracle_value),
               colour = "grey20", size = 1.8),
    geom_vline(xintercept = as.numeric(ref_date),
               linetype = "dashed", colour = "grey50"),
    scale_colour_manual(values = model_colors, name = NULL,
                        aesthetics = c("colour", "fill")),
    theme_minimal(base_size = 13),
    theme(legend.position = "top",
          plot.subtitle = element_text(face = "italic", colour = "grey30"))
  )

  median_layers <- list(
    geom_line(data = bands,
              aes(x = target_end_date, y = median, colour = model_id),
              linewidth = 1),
    geom_point(data = bands,
               aes(x = target_end_date, y = median, colour = model_id),
               size = 2.2)
  )

  trajectory_layer <- geom_line(
    data = forecast_display,
    aes(x = target_end_date, y = value,
        group = interaction(model_id, sample_id),
        colour = model_id),
    alpha = 0.12, linewidth = 0.3
  )

  if (scale == "linear") {
    y_scale_layer <- coord_cartesian(ylim = y_lim)
    y_label <- "% of ED visits / ILI"
    scale_tag <- "linear y"
  } else {
    y_scale_layer <- scale_y_log10(
      limits = y_lim_log, breaks = c(1, 2, 5, 10, 20),
      oob = scales::squish
    )
    y_label <- "% of ED visits / ILI (log10)"
    scale_tag <- "log y"
  }

  # Top panel: trajectories + median (carries the legend)
  top_plot <- ggplot() +
    base_layers + trajectory_layer + median_layers + y_scale_layer +
    labs(x = NULL, y = y_label) +
    theme(axis.text.x = element_blank(),
          plot.title = element_text(face = "bold"))

  # Bottom panel: 95% PI ribbon + median
  bottom_plot <- ggplot() +
    base_layers +
    geom_ribbon(data = bands,
                aes(x = target_end_date, ymin = lower, ymax = upper,
                    fill = model_id),
                alpha = 0.25) +
    median_layers + y_scale_layer +
    labs(x = "Target end date", y = y_label) +
    theme(legend.position = "none")

  (top_plot / bottom_plot) +
    plot_layout(heights = c(1, 1)) +
    plot_annotation(
      title = title,
      caption = sprintf("Reference date: %s (vertical dashed).", format(ref_date, "%Y-%m-%d")),
      theme = theme(
        plot.title = element_text(face = "bold", size = 14),
        plot.caption = element_text(colour = "grey40", size = 9)
      )
    )
}

# Score each forecaster on five scoring rules. Returns one row per model_id.
# All five metrics are returned in loss form (lower is better), matching the
# scoringutils conventions: `logs_sample` returns -log(estimated density at obs).
score_forecasts <- function(forecast_df, oracle_df) {
  models <- unique(forecast_df$model_id)
  bind_rows(lapply(models, function(m) {
    fd <- forecast_df |>
      filter(model_id == m) |>
      arrange(sample_id, target_end_date)
    horizons <- sort(unique(fd$target_end_date))
    # samples_per_horizon: rows = horizons, cols = samples
    spm <- fd |>
      select(sample_id, target_end_date, value) |>
      pivot_wider(names_from = sample_id, values_from = value) |>
      arrange(target_end_date) |>
      select(-target_end_date) |>
      as.matrix()
    obs <- oracle_df |>
      filter(target_end_date %in% horizons) |>
      arrange(target_end_date) |>
      pull(oracle_value)
    # log-scale transform (log1p handles 0 if any)
    spm_log <- log1p(spm)
    obs_log <- log1p(obs)
    # scoringutils convention for multivariate sample scores:
    # observed = length-d vector, predicted = d x N matrix, mv_group_id of length d.
    # All 4 horizons belong to ONE multivariate observation, so mv_group_id is all 1s.
    mv_id <- rep(1L, length(obs))
    tibble(
      model_id     = m,
      `CRPS`       = mean(crps_sample(observed = obs, predicted = spm)),
      `CRPS (log)` = mean(crps_sample(observed = obs_log, predicted = spm_log)),
      `LS`         = mean(logs_sample(observed = obs, predicted = spm)),
      `ES`         = energy_score_multivariate(observed = obs, predicted = spm, mv_group_id = mv_id),
      `VS`         = variogram_score_multivariate(observed = obs, predicted = spm, mv_group_id = mv_id, p = 0.5)
    )
  }))
}

# Helper to render the score table on an evaluation slide.
# - Model labels are coloured to match the figure.
# - Within each metric column the lower (better) value is shown in bold + that
#   model's colour.
#
# The HTML is emitted inside a `{=html}` raw block so that pandoc / Quarto's
# table post-processor does not strip the inline `style` attributes.
score_table <- function(scores, model_colors) {
  metric_cols <- setdiff(names(scores), "model_id")

  bold_color <- function(value, hex) {
    sprintf('<strong style="color:%s;">%s</strong>', hex, value)
  }

  # Header row
  header_cells <- paste0(
    '<th style="text-align:left;padding:0 0.4em;white-space:nowrap;">Model</th>',
    paste0(sprintf('<th style="text-align:right;padding:0 0.4em;white-space:nowrap;">%s</th>', metric_cols),
           collapse = "")
  )

  # Data rows
  rows <- character(nrow(scores))
  for (j in seq_len(nrow(scores))) {
    m <- scores$model_id[j]
    hex <- model_colors[[m]]
    name_cell <- sprintf(
      '<td style="text-align:left;padding:0 0.4em;white-space:nowrap;">%s</td>',
      bold_color(m, hex)
    )
    metric_cells <- vapply(metric_cols, function(col) {
      vals <- scores[[col]]
      val_str <- formatC(vals[j], digits = 2, format = "f")
      if (j == which.min(vals)) {
        cell_content <- bold_color(val_str, hex)
      } else {
        cell_content <- val_str
      }
      sprintf('<td style="text-align:right;padding:0 0.4em;">%s</td>',
              cell_content)
    }, character(1))
    rows[j] <- paste0("<tr>", name_cell, paste(metric_cells, collapse = ""), "</tr>")
  }

  html <- paste0(
    '<table style="border-collapse:collapse;margin:0;">',
    '<thead><tr>', header_cells, '</tr></thead>',
    '<tbody>', paste(rows, collapse = ""), '</tbody>',
    '</table>'
  )

  # Emit wrapped in raw-HTML block so Quarto leaves it alone.
  cat("\n:::: {=html}\n", html, "\n::::\n", sep = "")
  invisible(NULL)
}
```

## {.unlisted}

```{r ex1-setup}
ex1_ref_date <- as.Date("2025-11-22")
ex1_target_dates <- as.Date(c("2025-11-29", "2025-12-06",
                              "2025-12-13", "2025-12-20"))

# Model X: slightly increasing, shifted lognormal (floor at 2.0, asymmetric).
# 95% PI covers truth at all 4 horizons with margin; lower 95% PI stays around 3.
ex1_forecast_X <- generate_forecast(
  "Model X",
  medians = c(4.0, 4.5, 5.0, 5.5),
  sds     = c(0.45, 0.50, 0.65, 0.70),    # log-scale sigma (above the floor)
  dist    = "lognormal",
  floor   = 2.0,
  target_dates = ex1_target_dates,
  seed = 101
)

# Model Y: more steeply increasing but all medians under truth, narrow normal.
# 95% PI does not cover truth at any of h=1..4.
ex1_forecast_Y <- generate_forecast(
  "Model Y",
  medians = c(4.5, 5.3, 7.0, 9.0),
  sds     = c(0.40, 0.40, 0.75, 0.80),
  dist    = "normal",
  target_dates = ex1_target_dates,
  seed = 202
)

ex1_forecasts <- bind_rows(ex1_forecast_X, ex1_forecast_Y)

# Truth window: ~5 weeks of pre-reference history through the post-peak decline
ex1_truth_window <- truth |>
  filter(target_end_date >= as.Date("2025-10-25"),
         target_end_date <= as.Date("2026-02-14"))
```

::: {.panel-tabset}

### Natural scale

```{r ex1-plot-lin, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex1_truth_window, ex1_forecasts, ex1_ref_date,
  title = "Example 1 — NYC ILI ED visits %, forecasts from 2025-11-22",
  subtitle = "Model X: wide intervals, slow rise. Model Y: narrow intervals, steeper rise but always under truth.",
  scale = "linear"
)
```

### Log scale

```{r ex1-plot-log, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex1_truth_window, ex1_forecasts, ex1_ref_date,
  title = "Example 1 — NYC ILI ED visits %, forecasts from 2025-11-22",
  subtitle = "Model X: wide intervals, slow rise. Model Y: narrow intervals, steeper rise but always under truth.",
  scale = "log"
)
```

:::

## {.unlisted}

```{r ex1a-setup}
# Same reference date and same Model X as Example 1. Model Y here is slightly
# more biased low than Example 1's Y -- medians shifted down to (4.0, 4.5,
# 5.5, 7.0). Y is still narrow and follows the rising shape, but the larger
# absolute bias makes Y lose on linear CRPS (and CRPS-log, LS, ES) while it
# still wins on the shape-only variogram score.
ex1a_ref_date <- ex1_ref_date
ex1a_target_dates <- ex1_target_dates

ex1a_forecast_X <- generate_forecast(
  "Model X",
  medians = c(4.0, 4.5, 5.0, 5.5),
  sds     = c(0.45, 0.50, 0.65, 0.70),
  dist    = "lognormal",
  floor   = 2.0,
  target_dates = ex1a_target_dates,
  seed = 101
)

ex1a_forecast_Y <- generate_forecast(
  "Model Y",
  medians = c(4.0, 4.5, 5.5, 7.0),
  sds     = c(0.40, 0.40, 0.75, 0.80),
  dist    = "normal",
  target_dates = ex1a_target_dates,
  seed = 202
)

ex1a_forecasts <- bind_rows(ex1a_forecast_X, ex1a_forecast_Y)
ex1a_truth_window <- ex1_truth_window
```

::: {.panel-tabset}

### Natural scale

```{r ex1a-plot-lin, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex1a_truth_window, ex1a_forecasts, ex1a_ref_date,
  title = "Example 1a — NYC ILI ED visits %, forecasts from 2025-11-22",
  subtitle = "Model X: same as Example 1 (wide, slow rise). Model Y: narrow, slightly more biased low than in Example 1.",
  scale = "linear"
)
```

### Log scale

```{r ex1a-plot-log, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex1a_truth_window, ex1a_forecasts, ex1a_ref_date,
  title = "Example 1a — NYC ILI ED visits %, forecasts from 2025-11-22",
  subtitle = "Model X: same as Example 1 (wide, slow rise). Model Y: narrow, slightly more biased low than in Example 1.",
  scale = "log"
)
```

:::

## {.unlisted}

```{r ex2-setup}
# Reference date moved forward two weeks from Example 1.
ex2_ref_date <- as.Date("2025-12-06")
ex2_target_dates <- as.Date(c("2025-12-13", "2025-12-20",
                              "2025-12-27", "2026-01-03"))

# Model A: wide intervals, a small peak at h=2 then drifting down.
# 95% PI covers truth at all 4 horizons (including the peak).
ex2_forecast_A <- generate_forecast(
  "Model A",
  medians = c(6.0, 6.5, 5.0, 4.5),
  sds     = c(0.45, 0.60, 0.60, 0.50),    # log-scale sigma (above the floor)
  dist    = "lognormal",
  floor   = 2.0,
  target_dates = ex2_target_dates,
  seed = 303
)

# Model B: biased high; a soft peak at h=2 with a flatter decline (less of a
# clean mirror of the truth's rise-and-fall). Narrow normal: 95% PI does not
# cover any horizon.
ex2_forecast_B <- generate_forecast(
  "Model B",
  medians = c(10.5, 14.0, 13.5, 11.5),
  sds     = c(0.55, 0.85, 0.85, 0.70),
  dist    = "normal",
  target_dates = ex2_target_dates,
  seed = 404
)

ex2_forecasts <- bind_rows(ex2_forecast_A, ex2_forecast_B)

# Truth window: same span as Example 1 so the curve looks identical.
ex2_truth_window <- truth |>
  filter(target_end_date >= as.Date("2025-10-25"),
         target_end_date <= as.Date("2026-02-14"))
```

::: {.panel-tabset}

### Natural scale

```{r ex2-plot-lin, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex2_truth_window, ex2_forecasts, ex2_ref_date,
  title = "Example 2 — NYC ILI ED visits %, forecasts from 2025-12-06",
  subtitle = "Model A: wide intervals, drifting down. Model B: narrow intervals, right trend but overshoots the peak.",
  model_colors = c("Model A" = "#44AA99", "Model B" = "#AA4499"),
  y_lim = c(0, 18),
  scale = "linear"
)
```

### Log scale

```{r ex2-plot-log, fig.width=11, fig.height=6.5}
plot_forecast_panel(
  ex2_truth_window, ex2_forecasts, ex2_ref_date,
  title = "Example 2 — NYC ILI ED visits %, forecasts from 2025-12-06",
  subtitle = "Model A: wide intervals, drifting down. Model B: narrow intervals, right trend but overshoots the peak.",
  model_colors = c("Model A" = "#44AA99", "Model B" = "#AA4499"),
  y_lim = c(0, 18),
  scale = "log"
)
```

:::

## Example 1 — scores {.smaller}

::: {.columns}
::: {.column width="58%"}

::: {.panel-tabset}

#### Natural

```{r ex1-eval-plot-lin, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex1_truth_window, ex1_forecasts, ex1_ref_date,
  title = "Example 1 — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-11-22",
  scale = "linear"
)
```

#### Log

```{r ex1-eval-plot-log, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex1_truth_window, ex1_forecasts, ex1_ref_date,
  title = "Example 1 — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-11-22",
  scale = "log"
)
```

:::

:::
::: {.column width="42%"}

::: {style="font-size: 0.65em;"}

```{r ex1-eval-table}
#| results: asis
score_forecasts(ex1_forecasts, truth) |>
  score_table(model_colors = c("Model X" = "#4477AA",
                               "Model Y" = "#DDAA33"))
```

:::

*Lower is better for all five scores (log score is reported as the loss `-log(f(y))`).*

:::
:::

## Example 1a — scores {.smaller}

::: {.columns}
::: {.column width="58%"}

::: {.panel-tabset}

#### Natural

```{r ex1a-eval-plot-lin, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex1a_truth_window, ex1a_forecasts, ex1a_ref_date,
  title = "Example 1a — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-11-22 (Y nudged lower)",
  scale = "linear"
)
```

#### Log

```{r ex1a-eval-plot-log, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex1a_truth_window, ex1a_forecasts, ex1a_ref_date,
  title = "Example 1a — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-11-22 (Y nudged lower)",
  scale = "log"
)
```

:::

:::
::: {.column width="42%"}

::: {style="font-size: 0.65em;"}

```{r ex1a-eval-table}
#| results: asis
score_forecasts(ex1a_forecasts, truth) |>
  score_table(model_colors = c("Model X" = "#4477AA",
                               "Model Y" = "#DDAA33"))
```

:::

*Lower is better for all five scores (log score is reported as the loss `-log(f(y))`).*

:::
:::

## Example 2 — scores {.smaller}

::: {.columns}
::: {.column width="58%"}

::: {.panel-tabset}

#### Natural

```{r ex2-eval-plot-lin, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex2_truth_window, ex2_forecasts, ex2_ref_date,
  title = "Example 2 — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-12-06",
  model_colors = c("Model A" = "#44AA99", "Model B" = "#AA4499"),
  y_lim = c(0, 18),
  scale = "linear"
)
```

#### Log

```{r ex2-eval-plot-log, fig.width=7, fig.height=5}
plot_forecast_panel(
  ex2_truth_window, ex2_forecasts, ex2_ref_date,
  title = "Example 2 — NYC ILI ED visits %",
  subtitle = "Forecasts from 2025-12-06",
  model_colors = c("Model A" = "#44AA99", "Model B" = "#AA4499"),
  y_lim = c(0, 18),
  scale = "log"
)
```

:::

:::
::: {.column width="42%"}

::: {style="font-size: 0.65em;"}

```{r ex2-eval-table}
#| results: asis
score_forecasts(ex2_forecasts, truth) |>
  score_table(model_colors = c("Model A" = "#44AA99",
                               "Model B" = "#AA4499"))
```

:::

*Lower is better for all five scores (log score is reported as the loss `-log(f(y))`).*

:::
:::
